---
layout: post
title: "Modifying data with lookup tables"
author: Daniel Falster
date: 2013-05-14 08:20
comments: true
publish: true
categories: 
---

<!-- The problem:
- importing new data
- amount of code to be written (opportunities for mistake)
- separating data from scripts
- maintaining record of where data came from

Common approach
- long sequence of data modifying code

Solution
- use lookup table, find and replace
 -->


In many analyses, data is read from a file, but must be modified before it can be used. For example you may want to add a new column of data, or do a "find" and "replace" on a site, treatment or species name. There are 3 ways one might add such information. The first involves editing the original data frame -- although you should *never* do this, I suspect this method is quite common. A second -- and widely used -- approach for adding information is to modify the values using code in your script. The third -- and nicest -- way of adding information is to use a lookup table. 

<!-- more -->

One of the most common things we see in the code of researchers working with data are long slabs of code modifying a data frame based on some logical tests. Such code might correct a species name:


```
raw$species[raw$id=="1"] <- "Banksia oblongifolia"  
raw$species[raw$id=="2"] <- "Banksia ericifolia"
```

or add some details, for example in this case location, latitude, longitude and mean annual precipitation:

```
raw$location[raw$id=="1"] <-"NSW"
raw$latitude[raw$id=="1"] <- -37
raw$longitude[raw$id=="1"] <- 40
raw$map[raw$id=="1"] <- 1208
raw$map[raw$id=="1"] <- 1226
```

In large analyses, this type of code may go for hundreds of lines. 

Now before we go on, let me say that this approach to adding data is *much* better than editing your datafile directly, for the following two reasons:

1. It maintains the integrity of your raw data file
2. You can see where the new value came from (it was added in a script), and modify it later if needed.

There is also nothing *wrong* with adding data this way. However, it is what we would consider *messy* code, for these reaosns: 

- Long chunks of code modifying data is inherently difficult to read.
- There's a lot of typing involved, so lot's of work, and thus opportunities for error.
- It's harder to change variable names when they are embedded in code all over the place.

A far *nicer* way to add data to an existing data frame is to use a lookup table. Here is an exapmple of such a table, achieving similar (but not identical) modifications to the code above:


```{r, echo=FALSE}

# Generate data
generateData <- function(length=20){
  x <- runif(length)
  data <- data.frame(x=x, y = 2+ 3*x + runif(length(x)))
  data$id <- sort(sample(4, length(x), replace=TRUE)) 
  data$y <- data$y + data$id*0.2
  data
}

myData <- generateData()

#Load addNewData function from github
library(devtools, quietly=TRUE)
# source_gist("5589956") - broken in latest devtools, issue submitted

library(RCurl, quietly=TRUE)
if ( !file.exists("addNewData.r") ){
  url <- "https://gist.github.com/dfalster/5589956/raw/44c20bfcbbf03fcb227c7501d9afd3f32ec7215d/addNewData.R"
 cat(getURL(url, ssl.verifypeer = FALSE), file= "addNewData.r")  
  }

#download data file from github gist 5589956
library(RCurl, quietly=TRUE)
if ( !file.exists("dataNew.csv") ){
  url <- "https://gist.github.com/dfalster/5589956/raw/64c2ad2f58bc281538d52cf55671b580ad9b10af/dataNew.csv"
 cat(getURL(url, ssl.verifypeer = FALSE), file= "dataNew.csv")    
  }

```

```{r}
read.csv("dataNew.csv")
```


The columns of this table are

- **lookupVariable** is the name of the variable in the parent data we want to match against. If left  blank, change all rows.
- **lookupValue** is the value of lookupVariable to match against
- **newVariable** is the variable to be changed
- **newValue** is the value of `newVariable` for matched rows
- **source** includes any notes about where the data came from (e.g., who made the change)

So the table documents the changes we want to make to our dataframe. The function [addNewData.R](https://gist.github.com/dfalster/5589956) then takes this filename for this table and applies it to thedata frame. For example let's assume we have a data frame called `data`


```{r}
myData

```

and want to apply the table given above, we simply write

```{r}
source("addNewData.r")
allowedVars<-c("species", "family", "location")
addNewData("dataNew.csv", myData, allowedVars)
```

The large block of code is now reduced to a single line that clearly expresses what we want to achieve. Moreover, the new values (data) are stored as a table of *data* in a file, which is preferable to having data mixed in with our code. 

You can use this approach 
You can find the example files used here, as a [github gist](https://gist.github.com/dfalster/5589956)

```{r, echo=FALSE}

#cleanup 
d<-file.remove("dataNew.csv")
d<-file.remove("addNewData.r")

```

